Appendix

### 2025-12-19 manuscript_rmarkdown/manuscript.tex
Update manuscript for revision 3.

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-29}

@@

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that the sequence $\mathbf{X}$ was sampled
from a population that is distributed according to a central normal distribution.

The alternative hypothesis is that the sequence $\mathbf{X}$ was sampled from a
population that is not distributed according to a central normal distribution.
\end{definition}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 100000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulation_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

We introduced an empirical test for central normality to assess whether
sequences deviate from normality in a way that might require closer
inspection prior to further processing. The empirical central normality
test differs from other tests for normality, such as the Shapiro-Wilk
test \citep{Shapiro1965-zd}, as it assesses normality of the central
portion of a feature, instead of the entire feature. Compared to the
Shapiro-Wilk test, the empirical central normality test remains
consistent when a feature was not sampled from a normal distribution,
such as a central normal distribution with \(\kappa < 1.0\). This
enables assessing reasonable normality of (transformed) features in the
presence of outliers.

This work has the following limitation: We observed several numerical
stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed.

>>>>

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-12-19}

@@

If $\mathbf{X}$ is randomly sampled from a normal distribution,
$\sup_{r_j \in \mathbf{R}_{\text{c}}} r_j \rightarrow 0$, almost surely.
That is, every residual of the central portion of $\mathbf{X}$ will asymptotically
converge to 0.

The latter corollary follows from the definition of the residuals and
the restriction to normal distributions. Residuals were defined using
\(\frac{x_i - \mu_M}{\sigma_M} \approx \hat{F}^{-1}_{\mathcal{N}}(p_i)\),
with \(\hat{F}^{-1}_{\mathcal{N}}(p_i)\) the empirical quantile function
resembling the quantile function of the normal distribution
\(\mathcal{N}(0, 1)\). Since
\(\hat{F}^{-1}_{\mathcal{N}}(p) \rightarrow F^{-1}_{\mathcal{N}}(p)\)
with \(\frac{1-\kappa}{2} \leq  p \leq \frac{1 + \kappa}{2}\)
\citep{van-der-Vaart1998-xu},
\(\sup_{r_j \in \mathbf{R}_{\text{c}}} r_j \rightarrow 0\), if and only
if for all \(x_i \in \mathbf{X}_c\),
\(\frac{x_i - \mu_M}{\sigma_M} \rightarrow \hat{F}^{-1}_{\mathcal{N}}(p_i)\).
Thus, the requirement is that the estimators for location and scale
\(\mu_M\) and \(\sigma_M\) are bias-free. M-estimators for location and
scale satisfy this condition, if \(\mathbf{X}\) follows a normal
distribution without contamination {[}Huber1964-xk;
\citet{Huber1981-su}{]}.

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that $\mathbf{X}$ was randomly sampled
from a population that is distributed according to a central normal distribution.

The alternative hypothesis is that $\mathbf{X}$ was randomly sampled from a
population that is not distributed according to a central normal distribution.
\end{definition}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each feature. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each feature after transformation.
Then, each method was ranked according to the sum of residual errors
(data without outliers) or the sum of residual errors of the central
portion (data with outliers). The average rank of each transformation
method was computed over all 100000 features in each dataset. Average
ranks for Yeo-Johnson transformations are shown in Table
\ref{tab:comparison_methods_simulations_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

We introduced an empirical test for central normality to assess whether
features deviate from normality in a way that might require closer
inspection prior to further processing. The empirical central normality
test differs from other tests for normality, such as the Shapiro-Wilk
test \citep{Shapiro1965-zd}, as it assesses normality of the central
portion of a feature, instead of the entire feature. Compared to the
Shapiro-Wilk test, the empirical central normality test remains
consistent when a feature was not sampled from a normal distribution,
such as a central normal distribution with \(\kappa < 1.0\). This
enables assessing reasonable normality of (transformed) features in the
presence of outliers.

This work has the following limitations: We observed several numerical
stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Moreover,
the definition of central normal distribution does not constrain the
distribution of the non-central tails. Thus, the test may be
conservative for central normal distributions other than uncontaminated
normal distributions, because the location and scale estimators are not
generally bias-free \citep{Rousseeuw1994-ur}.

### 2025-12-19 manuscript_rmarkdown/manuscript_appendix.tex
Update appendix C
Update manuscript for revision 3.

\section{Appendix C: Central normality test and empirical central
normality test critical
values}\label{appendix-c-central-normality-test-and-empirical-central-normality-test-critical-values}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value. 
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Each sequence contained 30 samples or fewer. 
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30}
\begin{tabular}{l | l r r r r r r}

>>>>

\section{Appendix C: Empirical central normality test critical
values}\label{appendix-c}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. The clean dataset consists of features derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
If necessary, each feature was shifted so that every element had a strictly positive value. 
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 features each. Each feature contained 30 samples or fewer. 
The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30}
\begin{tabular}{l | l r r r r r r}
